---
title: "Generate Scope 3 disaggregation table (Tier 1/2/3+)"
author: "Damien Lieber @ DecarbNexus LLC"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---
<!--
BEGINNER NOTES
- You do not need to edit this file to use the outputs. Most users can just
  download the Excel/CSV in the outputs/ folder on GitHub.
- To reproduce the outputs locally: install R (and RStudio), open this file,
  and click the "Knit" button at the top in RStudio; or run in the Console:
    rmarkdown::render("analysis/generate_scope3_disaggregation_table_tier1_2_3+.Rmd")
- This document will:
  1) Install/load required R packages as needed
  2) Download the correct USEEIO model spec based on config.yml
  3) Build the model and compute Tier/Scope contributions
  4) Write Excel and CSV to outputs/
-->

```{r setup, include=FALSE}
# Global chunk options: hide code by default to keep the document readable.
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## 0. Read this first (for beginners)

- You can run this from RStudio with: `source("scripts/render.R")`.
- The results will be saved to the `outputs/` folder.
- Configuration lives in `config.yml` (SEF version and which sectors count as Scope 2).

## 1. Introduction

This R Markdown document performs a Structural Path Decomposition (SPD) on a specified USEEIO model. The goal is to disaggregate the total supply chain emission factors for each commodity into:

* **Tiers:** The position in the supply chain (Tier 1 = the commodity itself, Tier 2 = direct suppliers, Tier 3+ = all upstream suppliers).
* **Scopes:** The GHG accounting category of the emissions (Scope 1 or Scope 2).

The final outputs are an Excel workbook and a CSV, saved in the `outputs/` folder.

## 2. Setup & Configuration

This block handles all setup, including package installation and dynamically fetching model specifications from the official USEPA repository based on the user's selected `sef_version`.

```{r 01-setup-and-load-packages, echo=TRUE}
# --- 1. Package Management ---
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  yaml,       # For reading configuration files (local and remote)
  dplyr,      # For data manipulation
  reshape2,   # For the melt() function
  knitr,      # For creating nice tables
  httr,       # For checking URLs before trying to read them
  openxlsx,   # For writing to Excel files
  stringr,    # For advanced string manipulation
  tidyr       # For separate() function
)

# --- 2. Load User Configuration ---
config <- read_yaml("config.yml")
cat("User requested Supply Chain Factors (SEF) version:", config$sef_version, "\n")

# --- 3. Fetch Model Specifications from USEPA GitHub Repository ---
versioning_url <- "https://raw.githubusercontent.com/USEPA/supply-chain-factors/main/Versioning.yml"

# Check if the URL is accessible
response <- GET(versioning_url)
if (http_error(response)) {
  stop("Could not access the Versioning.yml file on GitHub. Please check your internet connection.")
}

# Read the versioning file directly from the URL's content
versioning_data <- read_yaml(text = content(response, "text", encoding = "UTF-8"))

# Look up the details for the user-specified SEF version
version_details <- versioning_data[[config$sef_version]]

# Error handling: Stop if the specified SEF version is not found
if (is.null(version_details)) {
  available_versions <- paste(names(versioning_data), collapse = ", ")
  stop(
    "The specified sef_version '", config$sef_version, "' was not found in the repository.\n",
    "Available versions are: ", available_versions
  )
}

# --- 4. Install and Load correct `useeior` version ---
useeior_tag <- versioning_data[[config$sef_version]][["useeior_tag"]]
useeior_ver <- versioning_data[[config$sef_version]][["useeior_ver"]]
installed_pkg <- installed.packages()
if (!"devtools"%in%installed_pkg[, "Package"]) {
  install.packages("devtools")
}
if (!"useeior"%in%installed_pkg[, "Package"]) {
  cli::cli_alert_info("Installing useeior v{useeior_ver} (tag @{useeior_tag}) from GitHub...")
  devtools::install_github(paste0("USEPA/useeior@", useeior_tag))
}
installed_useeior_ver <- installed_pkg[installed_pkg[, "Package"]=="useeior", "Version"]
if ("useeior"%in%installed_pkg[, "Package"] && useeior_ver!=installed_useeior_ver) {
  cli::cli_alert_warning(c("A new version of useeior (v{useeior_ver}) will be installed for generating SEF {SEF_version}. ",
                           "The useeior v{installed_useeior_ver} you have installed will be overwritten."))
  cli::cli_alert_info("Installing useeior v{useeior_ver} (tag @{useeior_tag}) from GitHub...")
  devtools::install_github(paste0("USEPA/useeior@", useeior_tag))
}

library(useeior)

cat("Setup complete. Ready to build the model.\n")
```

## 3. Build USEEIO Model

Now, we build the USEEIO model using the `model_name` and the corresponding remote model specification file fetched in the previous step.

```{r 02-build-model, cache=TRUE}
# --- 1. Prepare for local model specification file ---
spec_dir <- "spec_files"
# Create the directory if it doesn't exist
if (!dir.exists(spec_dir)) {
  dir.create(spec_dir)
}
# Define the path for the local specification file
model_name <- "USEEIOv2.2.22-GHG"
local_spec_file <- file.path(spec_dir, paste0(model_name, ".yml"))

# --- 2. Download the spec file if it doesn't exist locally ---
if (!file.exists(local_spec_file)) {

  model_spec_url <- paste0(
    "https://raw.githubusercontent.com/USEPA/supply-chain-factors/main/model-specs/",
    model_name,
    ".yml"
  )

  cat("Downloading model specification file from:", model_spec_url, "\n")
  response_spec <- GET(model_spec_url)
  
  # Stop with an error if the download fails
  if (http_error(response_spec)) {
    stop("Failed to download the model specification file: ", model_name, ".yml")
  }

  # Write the downloaded content to the local file
  writeBin(content(response_spec, "raw"), local_spec_file)
  cat("Model specification saved locally to:", local_spec_file, "\n")
} else {
  cat("Using existing local model specification file:", local_spec_file, "\n")
}

# --- 3. Build the model ---
# We now use the LOCAL file path for the configpaths argument.
# The `cache=TRUE` option will save time on subsequent runs.
model <- buildModel(model_name, configpaths = local_spec_file)
```

### Model Specifications

It's important to record the key specifications of the model used for this analysis to ensure reproducibility.

```{r 02a-model-specs}
# Extract key specifications from the model object
model_specs <- data.frame(
  Specification = c("USEEIO Supply Chain Emission Factors version", "Model Name", "Input-Output Model's Year", "GHG Data Year", "GHG Data File", "GHG Data File Location"),
  Value = c(
    config$sef_version,
    model$specs$Model,
    model$specs$IOYear,
    model$specs$SatelliteTable$GHG$DataYears,
    model$specs$SatelliteTable$GHG$StaticFile,
    model$specs$SatelliteTable$GHG$FileLocation
  )
)

# Print the specifications as a clean table
knitr::kable(model_specs, caption = "Key USEEIO Model Specifications")
```

## 4. Structural Path Decomposition (SPD)

This is the core of the analysis. We extract matrices from the model and partition them by scope *before* calculating the tier contributions.

```{r 03-spd-calculation, echo=TRUE}
# --- 1. Prepare Key Matrices ---
# D: Direct Impact Matrix, diagonally expanded from the vector in model$D
D_matrix <- diag(model$D[1, ])
colnames(D_matrix) <- colnames(model$D)
rownames(D_matrix) <- colnames(model$D)

# I: Identity Matrix (used for Tier 1 calculation)
I_matrix <- diag(nrow(D_matrix))
colnames(I_matrix) <- colnames(D_matrix)
rownames(I_matrix) <- rownames(D_matrix)

# A: Direct Requirements Matrix
A_matrix <- model$A

# A^2: squared Direct Requirements Matrix
A_squared_matrix <- A_matrix %*% A_matrix

# L: Leontief Inverse Matrix
L_matrix <- model$L

# --- 2. Prepare Scope-Specific Matrices ---
# Partition key matrices into Scope 1 and Scope 2 components
A_matrix_scope1 <- A_matrix
A_matrix_scope1[config$scope_2_sectors, ] <- 0

A_matrix_scope2 <- A_matrix
A_matrix_scope2[!(rownames(A_matrix_scope2) %in% config$scope_2_sectors), ] <- 0

# A^2: squared Direct Requirements Matrix
A_squared_matrix_scope2 <- A_squared_matrix
A_squared_matrix_scope2[!(rownames(A_squared_matrix_scope2) %in% config$scope_2_sectors), ] <- 0

# L: Leontief Inverse matrix partitions
L_matrix_scope1 <- L_matrix
L_matrix_scope1[config$scope_2_sectors, ] <- 0

L_matrix_scope2 <- L_matrix
L_matrix_scope2[!(rownames(L_matrix_scope2) %in% config$scope_2_sectors), ] <- 0

# --- 3. Calculate Total Impact for Each Scope ---
# This is used later to calculate the percentage contribution.
N_Total <- D_matrix %*% L_matrix
N_Total_scope1 <- D_matrix %*% L_matrix_scope1
N_Total_scope2 <- D_matrix %*% L_matrix_scope2

# --- 3.1. Tier 1 (Direct Impact: D * I) ---
N_Tier1_scope1 <- D_matrix %*% I_matrix
N_Tier1_scope2 <- D_matrix %*% A_matrix_scope2
N_Tier1 <- N_Tier1_scope1 + N_Tier1_scope2

# --- 3.2. Tier 2 (Direct Suppliers: D * A) ---
N_Tier2_scope1 <- D_matrix %*% A_matrix_scope1
N_Tier2_scope2 <- D_matrix %*% A_squared_matrix_scope2
N_Tier2 <- N_Tier2_scope1 + N_Tier2_scope2

# --- 3.3. Tier 3+ (Residual) ---
N_Tier3plus_scope1 <- N_Total_scope1 - N_Tier1_scope1 - N_Tier2_scope1
N_Tier3plus_scope2 <- N_Total_scope2 - N_Tier1_scope2 - N_Tier2_scope2
N_Tier3plus <- N_Total - N_Tier1 - N_Tier2
```

### Verification Step

Let's verify that our decomposition is correct. The sum of the tiers should equal the total impact matrix.

```{r 04-verification}
N_Check <- colSums(N_Tier1 + N_Tier2 + N_Tier3plus)

# Calculate the ratio of the summed tiers to the total impact for each commodity
Ratio_Check <- ifelse(
  abs(model$N) < 1e-12, 
  1,
  N_Check / model$N
)

# Check if all ratio values are close to 1 (within a tolerance for floating-point arithmetic)
verification_passed <- all(abs(Ratio_Check - 1) < 1e-9)
cat(paste("Disaggregation verification successful:", verification_passed, "\n"))
if (!verification_passed) {
  warning("Verification failed! The sum of tiers does not equal the total impact.")
}
```

## 5. Reshape Data, Prepare Tables, and Export to Excel

Now we process the matrix results into the final tables and export them to a multi-tab Excel file.

```{r 05-reshape-and-export, echo=TRUE}
# --- 1. Helper Function to Reshape Matrices ---
melt_and_label <- function(matrix, tier_label, scope_label) {
  df <- as.data.frame(matrix)
  df$Embedded_Sector_Code <- rownames(df)
  df_long <- reshape2::melt(
    df,
    id.vars = "Embedded_Sector_Code",
    variable.name = "Disaggregated_Commodity",
    value.name = "Absolute_Contribution_GHG"
  )
  df_long$Tier <- tier_label
  df_long$Scope <- scope_label
  # Keep only non-zero entries to reduce file size and improve readability
  df_long <- df_long[df_long$Absolute_Contribution_GHG != 0, ]
  return(df_long)
}

# --- 2. Reshape all Scope-Tier Matrices ---
df_t1_s1 <- melt_and_label(N_Tier1_scope1, "Tier 1", "Scope 1")
df_t1_s2 <- melt_and_label(N_Tier1_scope2, "Tier 1", "Scope 2")
df_t2_s1 <- melt_and_label(N_Tier2_scope1, "Tier 2", "Scope 1")
df_t2_s2 <- melt_and_label(N_Tier2_scope2, "Tier 2", "Scope 2")
df_t3plus_s1 <- melt_and_label(N_Tier3plus_scope1, "Tier 3+", "Scope 1")
df_t3plus_s2 <- melt_and_label(N_Tier3plus_scope2, "Tier 3+", "Scope 2")

# --- 3. Create the Main Contribution Table (by Code) ---
# --- 6. Clean Codes Before Export ---
abs_col_name <- paste0(
  "Absolute contribution (kgCO2e/USD_PRO_",
  model$specs$IOYear,
  " spent on embedded commodity)"
)
final_table_by_code <- bind_rows(
    df_t1_s1, df_t1_s2, df_t2_s1, df_t2_s2, df_t3plus_s1, df_t3plus_s2
  ) %>%
  mutate(
    Disaggregated_Commodity = stringr::str_replace(Disaggregated_Commodity, "/US$", ""),
    Embedded_Sector_Code    = stringr::str_replace(Embedded_Sector_Code, "/US$", "")
  ) %>%
  group_by(Disaggregated_Commodity) %>%
  mutate(
    Total_Commodity_GHG = sum(Absolute_Contribution_GHG),
    Relative_Contribution = (Absolute_Contribution_GHG / Total_Commodity_GHG)
  ) %>%
  ungroup() %>%
  select(
    Disaggregated_Commodity,
    Embedded_Sector_Code,
    Tier,
    Scope,
    Absolute_Contribution_GHG,
    Relative_Contribution
  ) %>%
  arrange(Disaggregated_Commodity, Tier, desc(Absolute_Contribution_GHG))

# --- 4. Create Supporting Tables for Excel Export ---

# Table with Names instead of Codes
industry_mapping <- model$Industries %>% select(Code, Name)

final_table_by_name <- final_table_by_code %>%
  left_join(industry_mapping, by = c("Disaggregated_Commodity" = "Code")) %>%
  rename(Disaggregated_Commodity_Name = Name) %>%
  left_join(industry_mapping, by = c("Embedded_Sector_Code" = "Code")) %>%
  rename(Embedded_Sector_Name = Name) %>%
  select(
    Disaggregated_Commodity,
    Disaggregated_Commodity_Name,
    Embedded_Sector_Code,
    Embedded_Sector_Name,
    Tier,
    Scope,
    Absolute_Contribution_GHG,
    Relative_Contribution
  )

# Table with Industry Details
sector_classification <- model$Industries %>%
  select(Code, `Industry name` = Name) %>%
  left_join(
    model$Commodities %>% select(Code, `Commodity name` = Name, Category, Subcategory, Description),
    by = "Code"
  ) %>%
  # Separate Category into Code and Name
  tidyr::separate(Category, into = c("Category Code", "Category Name"), sep = ": ", extra = "merge", remove = TRUE) %>%
  # Separate Subcategory into Code and Name
  tidyr::separate(Subcategory, into = c("Subcategory Code", "Subcategory Name"), sep = ": ", extra = "merge", remove = TRUE) %>%
  mutate(
    # Use a single regex to remove multiple possible prefixes
    Description = stringr::str_remove(Description, "^(BEA Code & Name is '?\\w+:.*?'?\\.\\s*)")
    ) %>%
  # Select and reorder the final columns for the Excel sheet
  select(
    `Category Code`,
    `Category Name`,
    `Subcategory Code`,
    `Subcategory Name`,
    `Sector code` = Code,
    `Sector name` = `Industry name`,
    `Commodity name`,
    Description
  )


# --- 5. Write all tables to a single Excel file ---
# Create a clean base filename from the model specs and ensure an outputs directory exists
outputs_dir <- "outputs"
if (!dir.exists(outputs_dir)) dir.create(outputs_dir, recursive = TRUE)

file_basename <- paste(
  "SEF",
  config$sef_version,
  "_disaggregation_factors_",
  paste0("GHG", model_specs[model_specs$Specification == "GHG Data Year", "Value"]),
  sep = "_",
  paste0("IO", model_specs[model_specs$Specification == "Input-Output Model's Year", "Value"])
)

excel_filename <- file.path(outputs_dir, paste0(file_basename, ".xlsx"))
csv_filename   <- file.path(outputs_dir, paste0(file_basename, ".csv"))

wb <- createWorkbook()

# --- Create and Add Author Info Tab (as the first sheet) ---
author_info <- data.frame(
  Field = c(
    "Author",
    "Organization",
    "Website",
    "Contact",
    "Open-source repository",
    "Q&A + Discussion"
  ),
  Value = c(
    "Damien Lieber",
    "DecarbNexus LLC",
    "decarbnexus.com",
    "contact@decarbnexus.com",
    "https://github.com/damienlieber-dnexus/useeio_sectors_disaggregation",
    "https://github.com/damienlieber-dnexus/useeio_sectors_disaggregation/discussions"
  ),
  stringsAsFactors = FALSE
)
addWorksheet(wb, "Author_Info")
writeData(wb, "Author_Info", author_info)

addWorksheet(wb, "Model_Specs")
writeData(wb, "Model_Specs", model_specs)

addWorksheet(wb, "Contributions_by_Name")
writeData(wb, "Contributions_by_Name", final_table_by_name %>%
  rename(!!abs_col_name := Absolute_Contribution_GHG))

addWorksheet(wb, "Contributions_by_Code")
writeData(wb, "Contributions_by_Code", final_table_by_code %>%
  rename(!!abs_col_name := Absolute_Contribution_GHG))

addWorksheet(wb, "sector_classification")
writeData(wb, "sector_classification", sector_classification)

addWorksheet(wb, "NAICS_to_USEEIO_crosswalk")
writeData(wb, "NAICS_to_USEEIO_crosswalk", model$crosswalk)

saveWorkbook(wb, excel_filename, overwrite = TRUE)

# Use the cleaned table for the CSV export
write.csv(final_table_by_code %>%
  rename(!!abs_col_name := Absolute_Contribution_GHG), csv_filename, row.names = FALSE, na = "")

cat("Saved Excel:", excel_filename, "\n")
cat("Saved CSV:  ", csv_filename, "\n")
```

## 6. Preview of Final Results

Here is a preview of the top 20 rows from the final contribution factor table (with codes). The full results, including the version with names, are in the Excel file.

```{r 06-preview-results}
# Using knitr::kable for a nicely formatted HTML table preview.
kable(
  head(final_table_by_name, 5),
  caption = "Top 5 Rows of the Final Contribution Factor Table (by Code)",
  digits = 6
)

```
